---
title: "Assignment HDS5101"
author: "Judith Beltrán López, 25241070"
date: "2026-03-05"
output: html_document
---


```{r}
## import ##
library(dplyr)
library(ggplot2)
library(ggcorrplot)
library(pROC)
library(CalibrationCurves)
library(glmnet)
library(e1071)
library(sjPlot)
library(MASS)
library(tree)
library(caret)

## load data ##
df <- read.csv("data/wdbc.data", header = FALSE)

colnames(df) <- c(
  "id", "diagnosis",
  "radius_mean", "texture_mean", "perimeter_mean", "area_mean", "smoothness_mean",
  "compactness_mean", "concavity_mean", "concave_points_mean", "symmetry_mean", "fractal_dimension_mean",
  "radius_se", "texture_se", "perimeter_se", "area_se", "smoothness_se",
  "compactness_se", "concavity_se", "concave_points_se", "symmetry_se", "fractal_dimension_se",
  "radius_worst", "texture_worst", "perimeter_worst", "area_worst", "smoothness_worst",
  "compactness_worst", "concavity_worst", "concave_points_worst", "symmetry_worst", "fractal_dimension_worst"
)
head(df)
```

```{r}
## AUXILIAR FUNCTIONS ##

brier_score <- function(y, pred_prob) {
  mean((y - pred_prob)^2)
}

compute_confusion_matrix <- function(y, pred_prob, thr=0.5){
  pred <- rep(0, length(pred_prob))
  pred[pred_prob>thr] <- 1
  table(pred, y)
}

misclassification_error <- function(cm){
  cm <- as.matrix(cm)
  1 - sum(diag(cm)) / sum(cm)
}
```

1. Goal of the prediction task

With the selected dataset (Breast Cancer Wisconsin), the goal is to develop and evaluate a prediction model that classifies a breast cancer mass as malignant or benign, using the different available predictors to estimate each probability of malignancy and produce a binary classification.

2. Description of the dataset (source, study design, population, size of data (nxp), missing values)

- *Source:* Breast Cancer Wisconsin, https://archive.ics.uci.edu/dataset/17/breast%2Bcancer%2Bwisconsin%2Bdiagnostic

- *Study design:* Observational cross-selectional dataset, each row is a single breast mass assessment based on measurements extracted from an FNA image.

- *Study population:* People with a suspicious breast mass who had a fine needle aspirate (FNA)

- *Size of data (nxp):* we have 569 observations and 32 columns, so our nxp will be 18208. But we have to exclude two columns, since one is our outcome and the other one is an ID column, which is not a predictor so the size of available data is 17070.

```{r}
size_of_data <- nrow(df)*ncol(df)
size_of_data

size_of_available_data <- nrow(df)*30
size_of_available_data
```

- *Missing values:* There are no missing values in this dataset.

3. A summary of the methods you use for the prediction task

*1. Logistic regression:* fit a standard binary logistic model to estimate the probability that the breast mass is malignant.

*2. LASSO:* fit a penalised logistic regression model with an L1 penalty to shrink coefficients and perform automatic variable selection; choose the penalty parameter using cross-validation.

*3. Naive Bayes:* use a probabilistic classifier for continuous predictors, modelling each predictor within each class and combining them under the conditional independence assumption.

*4. Linear Discrimination Analysis (LDA):* fit a generative classifier that assumes predictors are approximately normally distributed within each class and derives class probabilities using Bayes' theorem.

*5. Trees:* fit a classification tree to predict malignant vs benign by recursively splitting the predictor space into regions.

4. The results of building a prediction model or models, including evaluation of model or models, including evaluation of model performance

### Convert the outcome variable into binary:

Code the diagnosis column to binary column: 1=M, 0=B

```{r}
# delete id column because is just ID, we won't need it for the assignment
df <- subset(df, select = -c(id))

df$diagnosis <- ifelse(df$diagnosis=="M",1,0)
head(df)
```

### Split data into train/test:

```{r}
## Split data into train (70%)/ test (30%)
set.seed(1)

n <- nrow(df)
train <- runif(n) < 0.7

train_df <- df[train,]
test_df <- df[!train, ]
```

### Explore correlation between predictors:

```{r}
corr_mat <- cor(X_train)
corr_mat_plot <- ggcorrplot(
  corr_mat,
  type="upper",
  hc.order = TRUE,
  lab=FALSE
)

corr_mat_plot+theme(
  axis.text.x = element_text(size = 7, angle = 45, hjust = 1),
  axis.text.y = element_text(size = 7)
)

to_drop <- findCorrelation(corr_mat, cutoff = 0.90, names = TRUE)
to_drop
```

To reduce redundancy in train data, I will remove predictors with absolute pairwise correlation >0.9, based on the training data correlation matrix. So we need to update the train/test sets.

### Update train/test without found predictors

```{r}
train_df <- train_df[, !(names(train_df) %in% to_drop)]
test_df  <- test_df[,  !(names(test_df)  %in% to_drop)]

X_train <- subset(train_df, select=-diagnosis)
Y_train <- train_df$diagnosis
X_test <- subset(test_df, select=-diagnosis)
Y_test <- test_df$diagnosis
```

### Model 1: Logistic regression

```{r}
# 1. TRAIN MODEL
lr_model <- glm(
  diagnosis ~ .,
  data = train_df,
  family = binomial()
)

summary(lr_model)

# 2. PREDICT
lr_probs <- predict(lr_model, test_df, type = "response")

# 3. BRIER SCORE
lr_bs <- brier_score(Y_test, lr_probs)
lr_bs

# 4. CONFUSION MATRIX
lr_cm <- compute_confusion_matrix(Y_test, lr_probs)
lr_cm

# 5. ROC & AUC
lr_roc <- roc(Y_test, lr_probs)
plot(lr_roc, print.auc=TRUE)
auc(lr_roc)

# 6. CALIBRATION CURVE
lr_cc <- valProbggplot(lr_probs, Y_test)
lr_cc
```

### Model 2: LASSO

```{r}
# 1. TRAIN MODEL
X_train <- model.matrix(diagnosis~., train_df)[,-1]

set.seed(1)
cv_lasso <- cv.glmnet(
  x=X_train,
  y=Y_train,
  family="binomial",
  alpha=1 # <- 1 means LASSO
)

# 2. PREDICT
X_test <- model.matrix(diagnosis~., test_df)[,-1]
lasso_probs <- as.numeric(predict(
  cv_lasso,
  newx=X_test,
  s="lambda.min",
  type="response"
))

# 3. BRIER SCORE
lasso_bs <- brier_score(Y_test, lasso_probs)
lasso_bs

# 4. CONFUSION MATRIX
lasso_cm <- compute_confusion_matrix(Y_test, lasso_probs)
lasso_cm

# 5. ROC & AUC
lasso_roc <- roc(Y_test, lasso_probs)
plot(lasso_roc, print.auc=TRUE)
auc(lasso_roc)

# 6. CALIBRATION CURVE
lasso_cc <- valProbggplot(lasso_probs, Y_test)
lasso_cc
```

### Model 3: Naive Bayes

```{r}
# 1. TRAIN MODEL
nb_model <- naiveBayes(diagnosis~., data=train_df)

# 2. PREDICT
nb_class <- predict(nb_model, test_df)
nb_preds <- predict(nb_model, test_df, type="raw")

# 3. BRIER SCORE
nb_bs <- brier_score(Y_test, nb_preds)
nb_bs

# 4. CONFUSION MATRIX
nb_cm <- table(nb_class, Y_test)
nb_cm

# 5. ROC & AUC
nb_roc <- roc(Y_test, nb_preds[,2])
plot(nb_roc, print.auc=TRUE)
auc(nb_roc)

# 6. CALIBRATION CURVE
nb_cc <- valProbggplot(nb_preds[,2], Y_test)
nb_cc
```

### Model 4: LDA

```{r}
# 1. TRAIN MODEL
lda_model <- lda(diagnosis ~ ., data = train_df)

# 2. PREDICT
lda_pred <- predict(lda_model, test_df)
lda_probs <- lda_pred$posterior[,"1"]

# 3. BRIER SCORE
lda_bs <- brier_score(Y_test, lda_probs)
lda_bs

# 4. CONFUSION MATRIX
lda_cm <- compute_confusion_matrix(Y_test, lda_probs)
lda_cm

# 5. ROC & AUC
lda_roc <- roc(Y_test, lda_probs)
plot(lda_roc, print.auc=TRUE)
auc(lda_roc)

# 6. CALIBRATION CURVE
lda_cc <- valProbggplot(lda_probs, Y_test)
lda_cc
```

### Model 5: Trees

```{r}
train_tree <- train_df
test_tree  <- test_df

train_tree$diagnosis <- factor(train_tree$diagnosis, levels = c(0, 1))
test_tree$diagnosis  <- factor(test_tree$diagnosis,  levels = c(0, 1))

# 1. TRAIN MODEL
tree_model <- tree(diagnosis~., data=train_tree)

# 2. PREDICT
tree_pred <- predict(tree_model, newdata=test_tree, type="vector")
tree_probs <- tree_pred[,"1"]

# 3. BRIER SCORE
tree_bs <- brier_score(Y_test, tree_probs)
tree_bs

# 4. CONFUSION MATRIX
tree_cm <- compute_confusion_matrix(Y_test, tree_probs)
tree_cm

# 5. ROC & AUC
tree_roc <- roc(Y_test, tree_probs)
plot(tree_roc, print.auc=TRUE)
auc(tree_roc)

# 6. CALIBRATION CURVE
tree_cc <- valProbggplot(tree_probs, Y_test)
tree_cc
```

5. discussion/conclusions

```{r}
comparison_table <- data.frame(
  Model = c("Logistic Regression", "LASSO", "Naive Bayes", "LDA", "Tree"),

  Brier = c(
    lr_bs,
    lasso_bs,
    nb_bs,
    lda_bs,
    tree_bs
  ),

  MisclassificationError = c(
    misclassification_error(lr_cm),
    misclassification_error(lasso_cm),
    misclassification_error(nb_cm),
    misclassification_error(lda_cm),
    misclassification_error(tree_cm)
  ),

  AUC = c(
    as.numeric(auc(lr_roc)),
    as.numeric(auc(lasso_roc)),
    as.numeric(auc(nb_roc)),
    as.numeric(auc(lda_roc)),
    as.numeric(auc(tree_roc))
  ),

  CalibrationSlope = c(
    as.numeric(lr_cc$stats["Slope"]),
    as.numeric(lasso_cc$stats["Slope"]),
    as.numeric(nb_cc$stats["Slope"]),
    as.numeric(lda_cc$stats["Slope"]),
    as.numeric(tree_cc$stats["Slope"])
  )
)

comparison_table
```

### Test set performance comparison (errors, AUC, calibration)

* **Best overall model: LASSO**

  * Lowest **Brier score** (0.029) and highest **AUC** (0.995), indicating the best overall probability accuracy and discrimination.
  * Misclassification error is also low (~0.054), though not the lowest.

* **Lowest misclassification error: Logistic Regression**

  * Lowest **misclassification error** (0.048) at the 0.5 threshold.
  * However, it has a higher **Brier score** (0.048) and much lower **AUC** (0.949) than LASSO, suggesting weaker ranking/probability performance overall.

* **Naive Bayes: poor probability quality**

  * Very high **Brier score** (0.487) despite a reasonable **AUC** (0.946).
  * This pattern suggests the predicted probabilities are **poorly calibrated** (also reflected by a very low calibration slope ~0.16).

* **LDA: strong but not best**

  * High **AUC** (0.980) with moderate **Brier score** (0.049) and misclassification error (~0.065).
  * Competitive performance, but still dominated by LASSO.

* **Tree: acceptable but weaker**

  * Higher **Brier score** (0.061) and misclassification error (~0.071) with **AUC** similar to logistic regression (~0.948).
  * A single tree likely underfits compared with penalised/linear methods.

* **Calibration slopes**

  * A slope close to **1** is ideal.
  * Logistic regression (~0.08) and Naive Bayes (~0.16) indicate **substantial miscalibration**.
  * LASSO (~1.38) suggests calibration is improved but still not perfect; LDA (~0.69) and Tree (~0.84) are closer to 1 than LR/NB.


```{r}
# logistic regression
plot(lr_roc, type = "s", col="green")   

# LASSO
lines(lasso_roc,   type = "s", col = "red")

# Naive Bayes
lines(nb_roc, type = "s", col = "blue")  

# LDA
lines(lda_roc, type = "s", col = "orange")  

# Trees
lines(tree_roc, type = "s", col = "lightblue")  
```
### ROC curve comparison (test set)

* All five ROC curves sit well above the 45° reference line, indicating **good discrimination** for all models (they perform much better than chance).
* **LASSO (red)** shows the **best discrimination overall**, with the curve closest to the top-left corner across most thresholds (consistent with the highest AUC in your results).
* **Logistic regression (green)** is also very strong, but sits slightly below LASSO, suggesting **slightly weaker ranking** of malignant vs benign cases.
* **LDA (orange)** performs well and is close to the top performers, though generally not as consistently high as LASSO/logistic across the full range.
* **Tree (light blue)** is competitive but tends to be a bit lower than the linear/penalised methods in parts of the curve, consistent with slightly reduced discrimination.
* **Naive Bayes (blue)** is the weakest of the group on the ROC plot (lower sensitivity for the same specificity in several regions), although it still shows good discrimination overall.

Overall, the ROC plot supports the conclusion that **LASSO provides the best test-set discrimination**, with logistic regression and LDA close behind.
