---
title: "Assignment HDS5101"
author: "Judith Beltrán López, 25241070"
date: "2026-03-05"
output: html_document
---


```{r}
## import ##
library(dplyr)
library(ggplot2)
library(ggcorrplot)
library(pROC)
library(CalibrationCurves)
library(glmnet)
library(e1071)
library(sjPlot)

## load data ##
df <- read.csv("data/wdbc.data", header = FALSE)

colnames(df) <- c(
  "id", "diagnosis",
  "radius_mean", "texture_mean", "perimeter_mean", "area_mean", "smoothness_mean",
  "compactness_mean", "concavity_mean", "concave_points_mean", "symmetry_mean", "fractal_dimension_mean",
  "radius_se", "texture_se", "perimeter_se", "area_se", "smoothness_se",
  "compactness_se", "concavity_se", "concave_points_se", "symmetry_se", "fractal_dimension_se",
  "radius_worst", "texture_worst", "perimeter_worst", "area_worst", "smoothness_worst",
  "compactness_worst", "concavity_worst", "concave_points_worst", "symmetry_worst", "fractal_dimension_worst"
)
head(df)
```

```{r}
## AUXILIAR FUNCTIONS ##

brier_score <- function(y, pred_prob) {
  mean((y - pred_prob)^2)
}

compute_confusion_matrix <- function(y, pred_prob, thr=0.5){
  pred <- rep(0, length(pred_prob))
  pred[pred_prob>thr] <- 1
  table(pred, y)
}
```

1. Goal of the prediction task

With the selected dataset (Breast Cancer Wisconsin), the goal is to develop and evaluate a prediction model that classifies a breast cancer mass as malignant or benign, using the different available predictors to estimate each probability of malignancy and produce a binary classification.

2. Description of the dataset (source, study design, population, size of data (nxp), missing values)

- *Source:* Breast Cancer Wisconsin, https://archive.ics.uci.edu/dataset/17/breast%2Bcancer%2Bwisconsin%2Bdiagnostic

- *Study design:* Observational cross-selectional dataset, each row is a single breast mass assessment based on measurements extracted from an FNA image.

- *Study population:* People with a suspicious breast mass who had a fine needle aspirate (FNA)

- *Size of data (nxp):* we have 569 observations and 32 columns, so our nxp will be 18208. But we have to exclude two columns, since one is our outcome and the other one is an ID column, which is not a predictor so the size of available data is 17070.

```{r}
size_of_data <- nrow(df)*ncol(df)
size_of_data

size_of_available_data <- nrow(df)*30
size_of_available_data
```

- *Missing values:* There are no missing values in this dataset.

3. A summary of the methods you use for the prediction task

*1. Logistic regression:* fit a standard binary logistic model to estimate the probability that the breast mass is malignant.

*2. LASSO:* fit a penalised logistic regression model with an L1 penalty to shrink coefficients and perform automatic variable selection; choose the penalty parameter using cross-validation.

*3. Naive Bayes:* use a probabilistic classifier for continuous predictors, modelling each predictor within each class and combining them under the conditional independence assumption.

*4. Linear Discrimination Analysis (LDA):* fit a generative classifier that assumes predictors are approximately normally distributed within each class and derives class probabilities using Bayes' theorem.

*5. Trees:* fit a classification tree to predict malignant vs benign by recursively splitting the predictor space into regions.

4. The results of building a prediction model or models, including evaluation of model or models, including evaluation of model performance

### Convert the outcome variable into binary:

Code the diagnosis column to binary column: 1=M, 0=B

```{r}
# delete id column because is just ID, we won't need it for the assignment
df <- subset(df, select = -c(id))

df$diagnosis <- ifelse(df$diagnosis=="M",1,0)
head(df)
```

### Split data into train/test:

```{r}
## Split data into train (70%)/ test (30%)
set.seed(1)

n <- nrow(df)
train <- runif(n) < 0.7

train_df <- df[train,]
test_df <- df[!train, ]

Y_train <- train_df$diagnosis
Y_test <- test_df$diagnosis
```

### Model 1: Logistic regression

```{r}
# 1. TRAIN MODEL
lr_model <- glm(
  diagnosis ~ .,
  data = train_df,
  family = binomial()
)

summary(lr_model)

# 2. PREDICT
lr_probs <- predict(lr_model, test_df, type = "response")

# 3. BRIER SCORE
lr_bs <- brier_score(Y_test, lr_probs)
lr_bs

# 4. CONFUSION MATRIX
lr_cm <- compute_confusion_matrix(Y_test, lr_probs)
lr_cm

# 5. ROC & AUC
lr_roc <- roc(Y_test, lr_probs)
plot(lr_roc, print.auc=TRUE)
auc(lr_roc)

# 6. CALIBRATION CURVE
lr_cc <- valProbggplot(lr_probs, Y_test)
lr_cc
```

### Model 2: LASSO

```{r}
```

### Model 3: Naive Bayes

```{r}
```

### Model 4: LDA

```{r}
```

5. discussion/conclusions

